---
title: "Predicting disease risk from daily habits: a binary logistic regression analysis"
subtitle: A statistical analysis- Fall 2025
authors: "Kristina Kusem, Shree Basnet, Renan Barbosa (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

<!-- Formatting for background color of entire page -->
<style>
body {
  background-color: lightblue; /* Change to your desired color */
}
</style>

<!--Formatting for text color: \n 
<span style="color: midnightblue;">This text is blue.</span> And this text is <span style="color: blue;">blue.</span> -->


Slides: [slides.html](slides.html){target="_blank"} ( Go to `slides.qmd`
to edit)

<!--::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!

Nice report!
::: -->


<!-- The introduction should:

-   Develop a storyline that captures attention and maintains interest.

-   Your audience is your peers

-   Clearly state the problem or question you're addressing.



-   Introduce why it is relevant needs.

-   Provide an overview of your approach.

Example of writing including citing references:

*This is an introduction to ..... regression, which is a non-parametric
estimator that estimates the conditional expectation of two variables
which is random. The goal of a kernel regression is to discover the
non-linear relationship between two random variables. To discover the
non-linear relationship, kernel estimator or kernel smoothing is the
main method to estimate the curve for non-parametric statistics. In
kernel estimator, weight function is known as kernel function
[@efr2008]. Cite this paper [@bro2014principal]. The GEE [@wang2014].
The PCA [@daffertshofer2004pca]*. Topology can be used in machine
learning [@adams2021topology]

For Symbolic Regression [@wang2019symbolic] *This is my work and I want
to add more work...*

Cite new paper [@su2012linear]  -->

## Rough draft/ outline of Introduction and Literature Review:

### Introduction:

-State what logistic regression is and why it is commonly used in research (it helps make predictions about categorical outcome variables, usually a binary outcome. It also helps to analyze several predictors simultaneously and how they all relate to the outcome variable at once). Discuss how we usually model LR and how we interpret results.
-Discuss which types of scenarios logistic regression is usually used in. It is actually used in a variety of contexts and is very useful for solving everyday problems around the world:
 -disease detection- identifying predictors of the presence of a disease helps people make preventive decisions regarding their health. It also helps practitioners more easily identify and diagnose specific health conditions. 
-predictions about the way people make decisions; specifically, healthcare decisions
-traffic and accident prediction
-fraud detection

### Literature Review:

Articles about using logistic regression in medical contexts:

1. "Understanding logistic regression analysis" discusses the usefulness of a logistic regression model and describes how to set up and interpret such a model when analyzing data. A logistic regression model is especially useful when dealing with more than two predictors and an outcome variable that has two or more classes or categories. This article gives an example of a logistic regression analysis with a synthetic dataset about patients undergoing a drug treatment. The outcome variable is categorical and binary in nature, taking on values of survived (1), or did not survive (0). The result of the analysis gives an explanation of how to interpret output from the model. One must take the expnentials of the slopes in the model to find the chances (the probability) of an event occurring. It is noted that in order to correctly understand results of a logistic regression, one must carefully consider the differences between the odds ratio, the log odds, and the probabilities of events occurring. Another important point in the article states the importance of feature selection; a common way that predictors are selected for a logistic regression model is through a preliminary univariate analysis. During data preprocessing, all predictors are analyzed individually in relation to the outcome variable in a univariate analysis, and all significant predictors are used in the multivariate logistic regression analysis.  

2. "Determinants of coexistence of undernutrition and anemia among under- five children in Rwanda; evidence from 2019/20 demographic health survey: Application of bivariate binary logistic regression model"

-This article details a bivariate binary logistic regression model. There are two outcome variables: anemia and undernutrition in children under five years of age in Rwanda. The study analyzes the relationship between the two outcome variables, as well as the relationship between 26 predictors relating to the childrens’ preexisting health conditions, family information, details about the parents, and relevant geographic information. One result of the study was that the relationship between the two outcome variables, presence of malnutrition and presence of anemia, was found to be significant. It was also found that six predictors were found to be sinificant: mother’s age, drinking water, other children in household, child gender, birth order, and gender of head of household. The conclusion states that improving maternal education, supplementing with vitamin A and other nutrient dense foods, providing a healthy home environment environment, and decreasing maternal anemia may help improve rates of malnutrition and anemia in children.

3. Predictors of hospital admission when presenting with acute on chronic breathlessness: Binary logistic regression

4. Risk factors for airplane headache: A multivariate logistic regression analysis in a population of career flight personnel

5. Exploring the medical decision- making patterns and influencing factors among the general Chinese public: a binary logistic regression analysis


Articles about using logistic regression in traffic and engineering contexts:

1.	Binary logistic regression analysis of factors affecting urban road traffic safety.

2. Identification of factors influencing severity of motorcycle crashes in Dhaka,
Bangladesh using binary logistic regression model

Articles about detecting fraud:

1.	Using Binary logistic Regression to Detect Health Insurance Fraud


<!--## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```



```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References

-->
